---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r}
library(knitr)
library(ggplot2)
library(dplyr)
library(plyr)
```



```{r}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", temp, mode="wb")
unzip(temp, "activity.csv")
dataset <- read.csv("activity.csv",header=T)
unlink(temp)
```

## What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day

```{r}
totalstepsperday <- aggregate(steps ~ date, data = dataset, FUN = sum, na.rm = TRUE)
totalstepsperday
```

2. If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day

```{r}

# converting dates to Y-M-D format
dataset$date <- as.Date(dataset$date, "%Y-%m-%d")

png("TotalStepsPlot.png",width=480,height=480)
hist(totalstepsperday$steps, 
    main="Total Steps per Day", 
    xlab="Number of Steps per Day", 
    ylab = "Interval",
    col="blue",
    breaks=60)

dev.off()
```


3. Calculate and report the mean and median of the total number of steps taken per day

## Mean
```{r}
msteps <- mean(totalstepsperday$steps)
msteps
```

## Median 
```{r}
medsteps <- median(totalstepsperday$steps)
medsteps
```

## check work using summary
```{r}
summary(totalstepsperday)
```


## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
Fmins <- aggregate(steps ~ interval, data = dataset, FUN = mean, na.rm = TRUE)

png("Timeseries plot.png",width=480,height=480)
plot(x = Fmins$interval, 
    y = Fmins$steps, 
    type = "l", 
    col = "blue",
    xlab = "5-minute Timeline",
    ylab = "Average Steps Taken ~ Days",
    main = "Average Daily Activity Pattern")

dev.off()
```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
maxsteps <- Fmins$interval[which.max(Fmins$steps)]
maxsteps
```

## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
removemissedvalues <- sum(is.na(dataset$steps))
removemissedvalues
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
Replace NA values with the mean results for five minute intervals

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
newdataset <- dataset
nas <- is.na(newdataset$steps)
avg_interval <- tapply(newdataset$steps, newdataset$interval, mean, na.rm=TRUE, simplify = TRUE)
newdataset$steps[nas] <- avg_interval[as.character(newdataset$interval[nas])]
names(newdataset)
```
 Check for no-NA
```{r}
sum(is.na(newdataset))
```
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

Analysis without NAs:
```{r}
totalstepsperday2 <- aggregate(steps ~ date, data = newdataset, FUN = sum, na.rm = TRUE)
totalstepsperday2
```
Histogram without NA values
```{r}
png("Histogram without NAs.png",width=480,height=480)
hist(totalstepsperday2$steps, 
    main = "Total Steps per Day (no-NA)", 
    xlab = "Number of Steps per Day", 
    ylab = "Interval",
    col="blue",
    breaks=60)
dev.off()
```


## Are there differences in activity patterns between weekdays and weekends?
